# Reproducible Research: Peer Assessment 1

Daily walking activity data

## Loading and preprocessing the data

```{r, Load data, echo = TRUE}
unzip("./activity.zip") 
data <- read.csv("./activity.csv")
```

Let's see what's inside:
```{r Look into data, echo = TRUE}
head(data)
tail(data)
str(data)
summary(data)
```

How many missing values are there?
```{r Count NA-s, echo = TRUE}
sum(is.na(data))
sum(is.na(data$steps))
```
We see that all missing values are in "steps" column.  

## What is mean total number of steps taken per day?

Let's look at a simple a histogram.
```{r Histogram, echo = TRUE}
steps.per.day <- tapply(data$steps, data$date, sum)

library(ggplot2)
qplot(steps.per.day, geom="histogram", binwidth=diff(range(steps.per.day, na.rm=TRUE)) / 30) + geom_rug(color="blue", alpha=0.7) + labs(x="Steps per day", title="Histogram of total steps per day")
```

and at mean and median values:
```{r Mean and median, echo = TRUE}
mn <- mean(steps.per.day, na.rm=TRUE)
md <- median(steps.per.day, na.rm=TRUE)
mn
md
```

The mean number of total steps per day is **`r mn`** and the median is **`r md`**.


## What is the average daily activity pattern?

Daily activity pattern averaged within each five minute interval across all days:

```{r, fig.width=8, fig.height=4, echo = TRUE}
library(plyr)
# Split by intervals, find means
int.means <- ddply(data, "interval", summarise, mean=mean(steps, na.rm=TRUE))
ggplot(int.means, aes(interval, mean)) + geom_path() + labs(title="Average Nr of Steps in 5-minute Intervals", x="Interval", y="Average Nr of Steps")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r Most active interval, echo = TRUE}
idx <- which.max(int.means$mean)
int.means[idx,]
```

The interval with the maximum number of steps when averaged across all days in the dataset is **interval `r int.means[idx, "interval"]`**.

## Imputing missing values

Let's have a closer look at incomplete cases:

```{r Incomplete cases, echo = TRUE}
incomplete.cases <- sum(is.na(data$steps))
incomplete.cases
```
There are **`r incomplete.cases`** incomplete cases in the dataset. Let's look athow they are distributed in time.

```{r NA-s in time, echo = TRUE}
interval.na <- ddply(data, "interval", summarise, na.count=sum(is.na(steps)))
head(interval.na)
tail(interval.na)
```

It seems that there are always 8 missing values. Let's check:

```{r Calculate missing, echo = TRUE}
all(interval.na$na.count == 8)
```

Indeed! 
This probably means that there were 8 days with no data collected. 
To check this:

```{r NAs by day, echo = TRUE}
date.na <- ddply(data, "date", summarise, na.count=sum(is.na(steps)))
head(date.na)
sum(date.na$na.count != 0)
```

Which dates were these?

```{r Which days were these?, echo = TRUE}
dates.with.nas <- unique(date.na$date[date.na$na.count != 0])
as.vector(dates.with.nas)
```

We expect that activity patterns depend on weekday (less walking on weekends or vice versa)
Let's impute mkissing days' values by mean values of that weekday.

First, let's find wich day of week is the corresponding date.

```{r Add weekday, echo = TRUE}
data$weekday <- factor(weekdays(as.Date(data$date)))
head(data)
```

Which weekdays are these?

```{r Weekdays with NAs, echo = TRUE}
weekdays.with.na <- unique(data[data$date %in% dates.with.nas,]$weekday)
```

Interval means for each weekday

```{r Interval means for each weekday, echo = TRUE}
int.d.means <- ddply(data, c("interval", "weekday"), summarise, steps=mean(steps, na.rm=TRUE))
```

Now let's question whether or not computing the mean by weekday is going to be
any more informative than just substituting the overall mean across all days
(this is a bit of foreshadowing for the next question):

Fill in the missing values

```{r Fill in the missing values, echo = TRUE}
cmpl <- data[complete.cases(data),]
incmpl <- data[!complete.cases(data),]
merged <- merge(incmpl[, c("weekday", "interval", "date")], int.d.means, by=c("interval", "weekday"))
imputed <- rbind(cmpl, merged)
sum(!complete.cases(imputed)) # 0 
```

A histogram, mean and median of imputed data: 

```{r A histogram, mean and median of imputed data, fig.width=8, fig.height=4, echo = TRUE}
imp.steps.p.d <- tapply(imputed$steps, data$date, sum)
qplot(imp.steps.p.d, geom="histogram", binwidth=diff(range(imp.steps.p.d, na.rm=TRUE)) / 30) +
    geom_rug(color="blue", alpha=0.7) + labs(x="Steps per day", title="Histogram of total steps per day (for imputed data)")
imp.mn <- mean(imp.steps.p.d, na.rm=TRUE)
imp.md <- median(imp.steps.p.d, na.rm=TRUE)
imp.mn
imp.md
```

After imputation mean is **`r imp.mn`** and median is **`r imp.md`**.  
These are both different from the values before imputation: `r mn`, `r md`, respectively.


## Are there differences in activity patterns between weekdays and weekends?

Let's add a *weekend/weekday* boolean indicator to the data frame with imputed values:

```{r Add weekend indicator, fig.width=8, echo = TRUE}
imputed$day=ifelse(as.POSIXlt(as.Date(imputed$date))$wday%%6==0, "weekend","weekday")
imputed$day=factor(imputed$day,levels=c("weekday","weekend"))
stepsInterval2=aggregate(steps~interval+day,imputed,mean)
library(lattice)
xyplot(steps~interval|factor(day),data=stepsInterval2, aspect=1/2, type="l")
```

An obvious difference is that activity starts later on weekends.


